Clustering in a hyperbolic model of complex networks.

### 2019-02-27 introduction.tex
revised sections 6 and 7

As observed by Krioukov et al~\cite{krioukov2010hyperbolic}, and proved rigorously by Gugelmann et al.~\cite{gugelmann2012random}, the degree sequence of the KPKVB model follows a power-law with exponent $2\alpha+1$.
% More specifically, if $N_k$ denotes the number of vertices of degree exactly $k$ then Gugelmann et al.~\cite{gugelmann2012random} have shown that
% 
% \begin{equation}\label{eq:gugeldegseq} 
% N_k = (1+o(1)) \cdot n \cdot p_k \quad \text{ a.a.s., }
% \end{equation}
% 
% \noindent
% where $p_k := 2\alpha \cdot \xi^{2\alpha} \cdot \Gamma^+\left(k - 2\alpha, \xi\right) / k!$
% with, here and in the rest of the paper, $\xi := \frac{4\alpha\nu}{\pi(2\alpha-1)}$ and 
% $\Gamma^+(a,b) := \int_b^\infty t^{a-1} e^{-t} \text{d}t$ the 
% {\em upper incomplete gamma function} and, for $(E_n)_n$ a sequence of events, 
% $E_n$ {\em asymptotically almost surely} (a.a.s.) means that $\Pee(E_n)\to 1$.
%%In particular $X_n = (1+o(1)) Y_n$ is equivalent to $X_n/Y_n \plim 1$.
%In fact~\eqref{eq:gugeldegseq} was shown in~\cite{gugelmann2012random} to hold not only when $k$ is constant but 
%also if $k$ depends on $n$ as long as $k = o( n^\delta )$ for a suitable small constant $\delta = \delta( \alpha )$.
%Standard asymptotic manipulations show that 
%
%$$p_k \sim 2\alpha\xi^{2\alpha} k^{-(2\alpha+1)}, $$
%
%as $k\to\infty$.
%\TM{ \RD{We need to add something like the following} : 
%In Appendix~\ref{} we settle the minor technical point of showing that~\eqref{eq:gugeldegseq} in fact extends 
%to all $k$ with $k \ll n^{1/(2\alpha+1)}$, while 
%$N_k = 0$ a.a.s.~whenever $k \gg n^{1/(2\alpha+1)}$. }
Gugelmann et al~\cite{gugelmann2012random} also showed that the average degree converges in probability to the constant $8\nu\alpha^2/ \pi (2\alpha-1)^2$, and they showed that the (local) clustering coefficient is non-vanishing in the sense that it is bounded below by a positive constant a.a.s. Here, and in the rest of the paper, for a sequence $(E_n)_n$ of events, $E_n$ {\em asymptotically almost surely} (a.a.s.) means that $\Pee(E_n)\to 1$.
%Their work however left open the question of whether the clustering coefficient actually converges in probability.

@@

Earlier work of the first and third authors with Bode~\cite{bode2015largest} and of the first and third authors~\cite{fountoulakis2018law} has 
established the ``threshold for a giant component'': when $\alpha < 1$ then there always is a unique component of size linear in $n$ no matter how small 
$\nu$ (and hence the average degree) is; when $\alpha > 1$ all components are sublinear no matter the value of $\nu$; and when $\alpha=1$ then 
there is a critical value $\nu_{\text{c}}$ such that for $\nu < \nu_{\text{c}}$ all components are sublinear and for $\nu > \nu_{\text{c}}$ there 
is a unique linear sized component (all of these statements holding a.a.s.). 
Whether or not there is a giant component when $\alpha=1$ and $\nu=\nu_{\text{c}}$ remains an open problem.
In~\cite{KiwiMit} an ~\cite{KiwiMit2017+}, Kiwi and Mitsche considered the size of the second largest component and
showed that when $\alpha \in (1/2, 1)$, a.a.s., the second largest component has polylogarithmic order with exponent 
$1/(\alpha -1/2)$.

@@

In the literature there are unfortunately two distinct, rival definitions of the {\em clustering coefficient}.
One of those, sometimes called the {\em global} clustering coefficient, is defined as 
three times the ratio the number of triangles to the number of paths of length two in the graph.
Results for this version of the clustering coefficient in the KPKVB model were obtained by Candellero and the first author~\cite{candellero2016clustering} and for the evolution of graphs on more general spaces with negative curvature by the first author in~\cite{F12}.

In this paper we will study the other notion of clustering, the one which is also considered by Krioukov et al.~\cite{krioukov2010hyperbolic} and
Gugelmann et al.~\cite{gugelmann2012random}. 
It is sometimes called the {\em local} clustering coefficient, although we should point out that Gugelmann et al.~actually
call it the global clustering coefficient in their paper.
For a graph $G$ and a vertex $v\in V(G)$ we define the clustering coefficient {\em of $v$} as:
\[
	c(v) := \left\{\begin{array}{cl}
		\displaystyle \frac{1}{{\text{deg}(v)\choose 2}} \sum_{u,w\sim v} 1_{\{uw \in E(G)\}} 
			& \text{ if $\text{deg}(v) \geq 2$, }\\
		& \\
        0 & \text{ otherwise,}
        \end{array}\right.
\]
where $E(G)$ denotes the edge set of $G$ and $\text{deg}(v)$ is the degree of vertex $v$. That is, provided $v$ has degree at least two, $c(v)$ equals the number of edges that are actually present between the neighbours \PvdH{I think it is neighbors in US English vs neighbours in UK English. Is there any specific reason to us the UK version?} of $v$ divided by the
number of edges that could possibly be present between the neighbours given the degree of $v$.
The clustering coefficient of $G$ is now defined as the average of $c(v)$ over all vertices $v$:

@@

$$ c(k; G) := \begin{cases}
		\displaystyle \frac{1}{N_k} \sum_{v \in V(G), \atop \text{deg}(v)=k}  c(v)  &\mbox{ if } N_k \ge 1\\
		0 &\mbox{else.}
	\end{cases},
	$$

@@

We write $\Gamma(z) := \int_0^\infty t^{z-1} e^{-t}\text{d}t$ for the gamma function, 
$\Gamma^+(a,b) := \int_b^\infty t^{a-1} e^{-t}\text{d}t$ for the upper incomplete gamma function, 
 $B(a,b) := \int_0^1 u^{a-1}(1-u)^{b-1}\text{d}u = \Gamma(a)\Gamma(b) / \Gamma(a+b)$ for the beta function and 
 $B^-(x,a,b) := \int_0^x u^{a-1}(1-u)^{b-1}\text{d}u$ the lower incomplete beta function. 
We write $U(a,b,z)$ for the hypergeometric U-function (also called Tricomi's confluent hypergeometric function), which 
%for $a,b,z\in \mathbb{C}$, $b \not \in \mathbb{Z}_{\leq 0}$, $\mathrm{Re}(a), \mathrm{Re}(z) >0$ 
has the integral representation 
\[
	U(a,b,z) = \frac{1}{\Gamma(a)} \int_0^\infty e^{-zt} t^{a-1} (1+t)^{b-a-1} dt,
\] 
see~\cite[p.255 Equation (2)]{erdelyi1953higher}, and let $\MeijerGnew{m}{\ell}{p}{q}{{\bf a}}{{\bf b}}{z}$ denote 
Meijer's G-Function~\cite{meijer1946gfunction}, see Appendix~\ref{sec:Meijer_G_functions} for more details.

@@

\[
	c(k;G_n) \xrightarrow[n\to\infty]{\Pee} \gamma(k),
\]
where $\gamma(k)$ is defined for $\alpha \ne 1$ as 
\begin{align*}
\gamma(k)  =&\frac{1}{8\alpha (\alpha-1)\Gamma^+(k-2\alpha,\xi)} \left( -\Gamma^+(k - 2 \alpha, \xi) - 2\frac{\alpha (\alpha - 1/2)^2 \xi^{2} \Gamma^+(k - 2 \alpha - 2, \xi)}{(\alpha - 1)} \right. \\ 
&\left.+ 8 \alpha (\alpha - 1/2) \xi \Gamma^+(k - 2 \alpha - 1,\xi) \right.\\ 
&\left.+ 4\xi^{4\alpha - 2} \Gamma^+(k - 6 \alpha + 2, 
      \xi) \left( \frac{2^{ - 4\alpha}(3 \alpha - 1)}{(\alpha - 1)} + (\alpha - 1/2) B^-(1/2; 1 + 2 \alpha, -2 + 2 \alpha) \right)  \right.\\ 
&\left.+ \xi^{k-2\alpha} \Gamma(2\alpha+1)e^{-\xi} U(2\alpha+1,1+k-2\alpha,\xi) \right. \\ 
&\left.- \xi^{4\alpha-2} \Gamma(2\alpha+1)\MeijerG{3}{0}{2}{3}{1,3-2\alpha}{3-4\alpha,-6\alpha+k+2,0}{\xi}  \right)
\end{align*}
and for $\alpha = 1$ as the $\alpha\to1$ limit of the above expression.
% \begin{align*}
% 	c_\infty(k) &= \frac{9 \eta^3}{2 k!} 	
% 		\Gamma^+(k-3,\eta)-\frac{\xi^4}{k!}\frac{7+\pi^2}{4}\Gamma^+(k-4,\eta)\\
% 	&\hspace{10pt}+ \frac{\eta^k}{2k!}\int_0^1 (1-4z+3z^2)\ln(1-z)z^{1-k}e^{-\eta/z}\dd z\\ 
% 	&\hspace{10pt}+ \frac{\eta^k}{k!}\int_0^1 z^{3-k} \Li_2(z) e^{-\eta/z} \dd z,
% \end{align*}
% with $\eta = 4\nu/\pi$ and $\Li_2(z) = \sum_{t = 1}^\infty z^t/t^2$, the dilogarithm function.
\end{theorem}

@@

$$ \gamma(k) = 
(1+o(1)) \cdot \left\{ \begin{array}{cl}
\frac{8\alpha \nu}{\pi\left(4\alpha - 3\right)} \cdot k^{-1} &\text{ if } \alpha > \frac{3}{4}, \\
\frac{6 \nu}{\pi} \cdot \frac{\log \log (k/\xi)}{k}& \text{ if } \alpha = \frac{3}{4},\\
 c_{\alpha} \cdot k^{2-4\alpha} & \text{ if } \frac12 < \alpha < \frac34, 
\end{array} \right.,
$$
where $c_{\alpha} := \left( \frac{3\alpha - 1}{2^{4\alpha+1}\alpha(\alpha-1)^2} 
	+ \frac{(\alpha - \frac{1}{2})B^-(\frac{1}{2},2\alpha + 1, 2\alpha - 2)}{2(\alpha - 1)\alpha} 
	- \frac{B(2\alpha, 3\alpha - 4)}{4(\alpha - 1)} \right)  \cdot \xi^{4\alpha-2}$.
\end{proposition}

>>>>

As observed by Krioukov et al.~\cite{krioukov2010hyperbolic}, and proved rigorously by Gugelmann et al.~\cite{gugelmann2012random}, the degree sequence of the KPKVB model follows a power-law with exponent $2\alpha+1$.
% More specifically, if $N_k$ denotes the number of vertices of degree exactly $k$ then Gugelmann et al.~\cite{gugelmann2012random} have shown that
% 
% \begin{equation}\label{eq:gugeldegseq} 
% N_k = (1+o(1)) \cdot n \cdot p_k \quad \text{ a.a.s., }
% \end{equation}
% 
% \noindent
% where $p_k := 2\alpha \cdot \xi^{2\alpha} \cdot \Gamma^+\left(k - 2\alpha, \xi\right) / k!$
% with, here and in the rest of the paper, $\xi := \frac{4\alpha\nu}{\pi(2\alpha-1)}$ and 
% $\Gamma^+(a,b) := \int_b^\infty t^{a-1} e^{-t} \text{d}t$ the 
% {\em upper incomplete gamma function} and, for $(E_n)_n$ a sequence of events, 
% $E_n$ {\em asymptotically almost surely} (a.a.s.) means that $\Pee(E_n)\to 1$.
%%In particular $X_n = (1+o(1)) Y_n$ is equivalent to $X_n/Y_n \plim 1$.
%In fact~\eqref{eq:gugeldegseq} was shown in~\cite{gugelmann2012random} to hold not only when $k$ is constant but 
%also if $k$ depends on $n$ as long as $k = o( n^\delta )$ for a suitable small constant $\delta = \delta( \alpha )$.
%Standard asymptotic manipulations show that 
%
%$$p_k \sim 2\alpha\xi^{2\alpha} k^{-(2\alpha+1)}, $$
%
%as $k\to\infty$.
%\TM{ \RD{We need to add something like the following} : 
%In Appendix~\ref{} we settle the minor technical point of showing that~\eqref{eq:gugeldegseq} in fact extends 
%to all $k$ with $k \ll n^{1/(2\alpha+1)}$, while 
%$N_k = 0$ a.a.s.~whenever $k \gg n^{1/(2\alpha+1)}$. }
Gugelmann et al.~\cite{gugelmann2012random} also showed that the average degree converges in probability to the constant $8\nu\alpha^2/ \pi (2\alpha-1)^2$, and they showed that the (local) clustering coefficient is non-vanishing in the sense that it is bounded below by a positive constant a.a.s. Here, and in the rest of the paper, for a sequence $(E_n)_n$ of events, $E_n$ {\em asymptotically almost surely} (a.a.s.) means that $\Pee(E_n)\to 1$.
%Their work however left open the question of whether the clustering coefficient actually converges in probability.

@@

Earlier work of the first and third authors with Bode~\cite{bode2015largest} and of the first and third authors~\cite{fountoulakis2018law} has 
established the ``threshold for a giant component'': if $\alpha < 1$ then there is a unique component of size linear in $n$ no matter how small 
$\nu$ (and hence the average degree); if $\alpha > 1$ all components are sublinear no matter the value of $\nu$; and if $\alpha=1$ then 
there is a critical value $\nu_{\text{c}}$ such that for $\nu < \nu_{\text{c}}$ all components are sublinear and for $\nu > \nu_{\text{c}}$ there 
is a unique linearly sized component (all of these statements holding a.a.s.). 
Whether or not there is a giant component if $\alpha=1$ and $\nu=\nu_{\text{c}}$ remains an open problem.
In~\cite{KiwiMit} and~\cite{KiwiMit2017+}, Kiwi and Mitsche considered the size of the second largest component and
showed that for $\alpha \in (1/2, 1)$, a.a.s., the second largest component has polylogarithmic order with exponent 
$1/(\alpha -1/2)$.

@@

In the literature there are unfortunately two distinct, rival definitions of the {\em clustering coefficient}.
One of those, sometimes called the {\em global} clustering coefficient, is defined as 
three times the ratio of the number of triangles to the number of paths of length two in the graph.
Results for this version of the clustering coefficient in the KPKVB model were obtained by Candellero and the first author~\cite{candellero2016clustering} and for the evolution of graphs on more general spaces with negative curvature by the first author in~\cite{F12}.

In this paper we will study the other notion of clustering, the one which is also considered by Krioukov et al.~\cite{krioukov2010hyperbolic} and
Gugelmann et al.~\cite{gugelmann2012random}. 
It is sometimes called the {\em local} clustering coefficient, although we should point out that Gugelmann et al.~actually
call it the global clustering coefficient in their paper.
For a graph $G$ and a vertex $v\in V(G)$ we define the clustering coefficient {\em of $v$} as:
\[
	c(v) := \left\{\begin{array}{cl}
		\displaystyle \frac{1}{{\text{deg}(v)\choose 2}} \sum_{u,w\sim v} 1_{\{uw \in E(G)\}}, 
			& \text{ if $\text{deg}(v) \geq 2$, }\\
		& \\
        0, & \text{ otherwise,}
        \end{array}\right.
\]
where $E(G)$ denotes the edge set of $G$ and $\text{deg}(v)$ is the degree of vertex $v$. That is, provided $v$ has degree at least two, $c(v)$ equals the number of edges that are actually present between the neighbours \PvdH{I think it is neighbors in US English vs neighbours in UK English. Is there any specific reason to us the UK version?} of $v$ divided by the
number of edges that could possibly be present between the neighbours given the degree of $v$.
The clustering coefficient of $G$ is now defined as the average of $c(v)$ over all vertices $v$:

@@

$$ c(k; G) := \begin{cases}
		\displaystyle \frac{1}{N_k} \sum_{v \in V(G), \atop \text{deg}(v)=k}  c(v),  &\mbox{ if } N_k \ge 1,\\
		0, &\mbox{else,}
	\end{cases}
	$$

@@

We write $\Gamma(z) := \int_0^\infty t^{z-1} e^{-t}\text{d}t$ for the gamma function, 
$\Gamma^+(a,b) := \int_b^\infty t^{a-1} e^{-t}\text{d}t$ for the upper incomplete gamma function, 
 $B(a,b) := \int_0^1 u^{a-1}(1-u)^{b-1}\text{d}u = \Gamma(a)\Gamma(b) / \Gamma(a+b)$ for the beta function and 
 $B^-(x,a,b) := \int_0^x u^{a-1}(1-u)^{b-1}\text{d}u$ for the lower incomplete beta function. 
We write $U(a,b,z)$ for the hypergeometric U-function (also called Tricomi's confluent hypergeometric function), which 
%for $a,b,z\in \mathbb{C}$, $b \not \in \mathbb{Z}_{\leq 0}$, $\mathrm{Re}(a), \mathrm{Re}(z) >0$ 
has the integral representation 
\[
	U(a,b,z) = \frac{1}{\Gamma(a)} \int_0^\infty e^{-zt} t^{a-1} (1+t)^{b-a-1} dt,
\] 
see~\cite[p.255 Equation (2)]{erdelyi1953higher}, and let $\MeijerGnew{m}{\ell}{p}{q}{{\bf a}}{{\bf b}}{z}$ denote 
Meijer's G-Function~\cite{meijer1946gfunction}, see Appendix~\ref{sec:Meijer_G_functions} for more details.

@@

\[
	c(k;G_n) \xrightarrow[n\to\infty]{\Pee} \gamma(k),
\]
where $\gamma(k)$ is defined for $\alpha \ne 1$ as 
\begin{align*}
\gamma(k)  =&\frac{1}{8\alpha (\alpha-1)\Gamma^+(k-2\alpha,\xi)} \left( -\Gamma^+(k - 2 \alpha, \xi) - 2\frac{\alpha (\alpha - 1/2)^2 \xi^{2} \Gamma^+(k - 2 \alpha - 2, \xi)}{(\alpha - 1)} \right. \\ 
&\left.+ 8 \alpha (\alpha - 1/2) \xi \Gamma^+(k - 2 \alpha - 1,\xi) \right.\\ 
&\left.+ 4\xi^{4\alpha - 2} \Gamma^+(k - 6 \alpha + 2, 
      \xi) \left( \frac{2^{ - 4\alpha}(3 \alpha - 1)}{(\alpha - 1)} + (\alpha - 1/2) B^-(1/2; 1 + 2 \alpha, -2 + 2 \alpha) \right)  \right.\\ 
&\left.+ \xi^{k-2\alpha} \Gamma(2\alpha+1)e^{-\xi} U(2\alpha+1,1+k-2\alpha,\xi) \right. \\ 
&\left.- \xi^{4\alpha-2} \Gamma(2\alpha+1)\MeijerGnew{3}{0}{2}{3}{1,3-2\alpha}{3-4\alpha,-6\alpha+k+2,0}{\xi}  \right)
\end{align*}
and for $\alpha = 1$ as the $\alpha\to1$ limit of the above expression.
% \begin{align*}
% 	c_\infty(k) &= \frac{9 \eta^3}{2 k!} 	
% 		\Gamma^+(k-3,\eta)-\frac{\xi^4}{k!}\frac{7+\pi^2}{4}\Gamma^+(k-4,\eta)\\
% 	&\hspace{10pt}+ \frac{\eta^k}{2k!}\int_0^1 (1-4z+3z^2)\ln(1-z)z^{1-k}e^{-\eta/z}\dd z\\ 
% 	&\hspace{10pt}+ \frac{\eta^k}{k!}\int_0^1 z^{3-k} \Li_2(z) e^{-\eta/z} \dd z,
% \end{align*}
% with $\eta = 4\nu/\pi$ and $\Li_2(z) = \sum_{t = 1}^\infty z^t/t^2$, the dilogarithm function.
\end{theorem}

@@

$$ \gamma(k) = 
(1+o(1)) \cdot \left\{ \begin{array}{cl}
\frac{8\alpha \nu}{\pi\left(4\alpha - 3\right)} \cdot k^{-1} &\text{ if } \alpha > \frac{3}{4}, \\
\frac{6 \nu}{\pi} \cdot \frac{\log(k)}{k}& \text{ if } \alpha = \frac{3}{4},\\
 c_{\alpha} \cdot k^{2-4\alpha} & \text{ if } \frac12 < \alpha < \frac34, 
\end{array} \right.,
$$
where $c_{\alpha} := \left( \frac{3\alpha - 1}{2^{4\alpha+1}\alpha(\alpha-1)^2} 
	+ \frac{(\alpha - \frac{1}{2})B^-(\frac{1}{2},2\alpha + 1, 2\alpha - 2)}{2(\alpha - 1)\alpha} 
	- \frac{B(2\alpha, 3\alpha - 4)}{4(\alpha - 1)} \right)  \cdot \xi^{4\alpha-2}$.
\end{proposition}

### 2019-02-27 proof_overview_finite_k.tex

\begin{equation}\label{eq:def_intensity_function_f}
	f(x,y) = \frac{\alpha \nu}{\pi} e^{-\alpha y} \cdot 1_{\{y>0\}}.
\end{equation}

@@

For any point $p \in \Rcal$, we write $\BallPo{p}$ to denote the \emph{ball} around $p$, i.e.

\begin{equation}\label{eq:def_ball_P}
	\BallPo{p} = \{p^\prime \in \eR\times[0,\infty) : |x - x^\prime| \leq e^{\frac{y + y^\prime}{2}}\}.
\end{equation}

@@

For the definition of the finite graph, recall that in the definition of the KPKVB model we set $R = 2\log(n/\nu)$.
We consider the box $\Rcal = (-\frac{\pi}{2}e^{R/2}, \frac{\pi}{2}e^{R/2}] \times (0, R]$ in $\eR^2$. 
Then the \emph{finite box model} $\Gbox := \Gbox(n;\alpha, \nu)$ has vertex set $\mathcal{V} := \mathcal{P} \cap \Rcal$ and edges set such that
\[
	pp' \in E(\Gbox(n;\alpha, \nu)) \iff |x - x'|_{\pi e^{R/2}} \leq e^{\frac{y + y'}{2}},
\]
where $|x|_{r} = \min( |x|, r - |x|)$ for $-r\leq x\leq r$. 
Using $|.|_{\pi e^{R/2}}$ instead of $|.|$ results in the left and right boundaries of the box $\Rcal$ getting 
identified, which in particular 
%excludes boundary effects and 
makes the model invariant under horizontal shifts and reflections in vertical lines. 
The graph $\Gbox$ can thus be seen as a subgraph of $\Ginf$ induced on $\mathcal{V}_n$, with some additional edges caused by 
the identification of the boundaries.

>>>>

\begin{equation}\label{eq:def_intensity_function_f}
	f(x,y) = \frac{\alpha \nu}{\pi} e^{-\alpha y} \cdot \1_{\{y>0\}}.
\end{equation}

@@

For any point $p \in \R \times (0,\infty)$, we write $\BallPo{p}$ to denote the \emph{ball} around $p$, i.e.

\begin{equation}\label{eq:def_ball_P}
	\BallPo{p} = \{p^\prime \in \eR\times(0,\infty) : |x - x^\prime| \leq e^{\frac{y + y^\prime}{2}}\}.
\end{equation}

@@

For the definition of the finite graph, recall that in the definition of the KPKVB model we set $R = 2\log(n/\nu)$.
We consider the box $\Rcal = (-\frac{\pi}{2}e^{R/2}, \frac{\pi}{2}e^{R/2}] \times (0, R]$ in $\eR^2$. 
Then the \emph{finite box model} $\Gbox := \Gbox(n;\alpha, \nu)$ has vertex set $\mathcal{V}_n := \mathcal{P} \cap \Rcal$ and edge set such that
\[
	pp' \in E(\Gbox(n;\alpha, \nu)) \iff |x - x'|_{\pi e^{R/2}} \leq e^{\frac{y + y'}{2}},
\]
where $|x|_{r} = \min( |x|, r - |x|)$ for $-r\leq x\leq r$. 
Using $|.|_{\pi e^{R/2}}$ instead of $|.|$ results in the left and right boundaries of the box $\Rcal$ getting 
identified, which in particular 
%excludes boundary effects and 
makes the model invariant under horizontal shifts and reflections in vertical lines. 
The graph $\Gbox$ can thus be seen as a subgraph of $\Ginf$ induced on $\mathcal{V}_n$, with some additional edges caused by 
the identification of the boundaries.
